---
title: 'Winning the battle, but losing the WAR: Examining the value of skater injuries
  in the NHL'
author: "Luke Steer (@lukeasteer)"
date: "April 25, 2019"
output:
  html_document:
    theme: default
    highlight: default
---

### All players are equal, but some players are more equal than others

Not all player injuries in the NHL are equivalent, which is a concept so intuitive that it hardly requires explaining. For example, you’d be hard-pressed to find an analyst or fan who believes losing Kyle Brodziak for four weeks is as impactful to the Edmonton Oilers as losing Connor McDavid. However, the discussion of injuries in the NHL has historically been centred around Man Games Lost (MGL), a basic statistic which limits the quality of analysis.

Analysts often reference MGL as a means of explaining a team's under-performance in a particular season, but rarely do they holistically evaluate which players were missing, and what value those players would have been expected to provide had they played. More modern statistics such as Cap Hit of Injured Player (CHIP)[^1] and Time-On-Ice of Injured Player (TOIIP) attempt to account for the value of the player that was lost to injury, but fall short, given that salary is influenced by player seniority and market factors that vary on a year-to-year basis, while time-on-ice is influenced by coaching staff bias and decision-making. 

Since the origin of hockey analytics, there have been attempts to develop single-value evaluation metrics, which are useful for comparison of total player value across different teams and seasons. Early examples include Ryder’s Player Contribution (2003)[^2], Awad’s Goals Versus Threshold (2009)[^3], MacDonald’s Adjusted Plus-Minus (2012)[^4], Schuckers and Curro’s Total Hockey Ratings (2013)[^5], Gramacy, Taddy, and Jensen’s Logistic Event Ratings (2013)[^6], and Thomas et al’s Mean Even Strength Hazard Model (2013)[^7]. 

Wins Above Replacement (WAR), first developed by sabermetricians in baseball circles, is an evaluative metric intended to approximate a player’s impact in units of wins added, measured against what a replacement level player would be expected to offer in the same circumstances[^8]. Replacement level is not uniquely defined, but it is commonly equated to players earning a league-minimum salary. Recently, this framework has been applied in the context of the NHL by Thomas et al. (2014)[^9], Sprigings (2016)[^10], Perry (2017)[^11], and Younggren and Younggren (2018)[^12]. 

Unsurprisingly, teams whose players produce more WAR in aggregate (team WAR) tend to have better records. Team WAR and team points percentage were strongly correlated for the seasons examined in this study (Adjusted R-squared = 0.6576). This finding is supported by Cam Lawrence's 2015 study[^13], which found that team strength, in terms of Spriging's Goals Above Replacement (a WAR sibling), was strongly correlated with team points for seasons between 2009-2010 and 2014-2015 (R-squared = 0.7231). As such, we can assume that losing a player who is highly valued by WAR would have a more significant impact on team performance than losing a player with lesser value. Clearly, player value in terms of WAR should be considered when evaluating injuries, given that differences in team WAR help explain differences in team performance. 

Furthermore, accounting for value lost to injury when examining injury trends enables rudimentary analysis of sports science staff efficacy. A team’s most productive players tend to play the most and are subjected to a higher degree of injury risk, both in terms of overuse injuries, and exposure to damaging on-ice events such as hits and blocked shots. However, it is common for elite players to receive special or preferential training and medical treatment that should, in theory, mitigate their injury risk to some extent. An efficient sports science team is one that is able to identify risk factors and reduce training load and exposure prior to injury, ultimately improving the player's availability, performance, and total output. A successful recovery and return to play post-injury is more strongly influenced by player physiology than specific treatment methods, but accurate injury identification is also within the realm of sports science staff responsibility. The more quickly an injury can be accurately diagnosed, the sooner the sports science staff can develop an appropriate recovery plan and plan the player's rehabilitation. 

Mistakes and misdiagnoses can be costly, for both the team and the player. The Pittsburgh Penguins infamously dismissed Dr. Charles Burke and the rest of their medical staff in July 2012 following the misdiagnosis of Sidney Crosby's concussion, and his ensuing health issues. While neither the Penguins nor Crosby have publicly linked the two events, it is difficult to believe that Burke's mistakes did not play a role in the Penguins' decision to move on from him, and develop a new sports science solution. In developing more nuanced metrics for evaluating value lost to injury, we are able to more accurately identify teams with recurring injury problems, which is the first step towards developing appropriate solutions to these problems.

### Method

9 seasons of data were examined, from the 2009-2010 NHL season through the 2017-2018 NHL season. For the purposes of this evaluation, Perry's WAR model was used as a measurement of player value, and data was sourced from corsica.hockey. Younggren and Younggren's WAR model was considered as an alternative to Perry's, and was preferable in that it encompasses regular season results exclusively. However, Younggren and Younggren's WAR is meant to be descriptive of a player's performance in a single season, whereas Perry's WAR emphasizes predictivity, and offers a better approximation of true player talent. As such, Perry's version of WAR was the preferred option for this analysis. Injury data, including CHIP, was sourced from nhlinjuryviz.blogspot.com. Team performance data was sourced from hockey-reference.com.

Assumptions:

* Perry does not differentiate between regular season games and playoff games when calculating WAR, all games are incorporated into a single value for the season. I’ve assumed that:
    + On aggregate, player performance is relatively stable between the regular season and the playoffs. 
    + WAR is more reflective of regular season performance due to the comparatively smaller number of playoff games in the sample. 
* The quality of the replacement player was not considered. My aim is not to reward deep organizations, nor to penalize shallow ones; this is strictly an injury evaluation.

Required packages and data sets were loaded, and the data was examined.

```{r warning=FALSE, message=FALSE}
# Load required packages
library(tidyverse)
library(ggthemes)
library(MASS)
library(robustbase)

# dplyr::select is masked by MASS::select, so specify the requirement
select <- dplyr::select

# Load WAR data, injury data, and team performance data 
war_df <- read_csv("war_ratings_2019-02-14.csv")
colnames(war_df)

injury_df <- read_csv("NHL_Injury_Database_data.csv")
colnames(injury_df)

team_stats_df <- read_csv("NHL_team_data.csv")
colnames(team_stats_df)
```

Required data was selected from the war_df data frame, and player trades and non-injury absences were accounted for. 

```{r warning=FALSE}
# Create a df with the required WAR data 
war82_df <- war_df %>%
  select(player = Player, 
         season = Season, 
         team = Team, 
         games_played = GP, 
         toi = TOI, 
         war = WAR, 
         war_82 = `WAR/82`, 
         war_60 = `WAR/60`) %>%
  # Account for mid-season player trades
  separate(team, c("team_one","team_two", "team_three", "team_four"), sep = '/') %>%
  # Create games_missed var, for comparison to games_missed_injury; account for non-injury absences 
  mutate(games_missed = 82 - games_played)
```

A weighted WAR was calculated in order to stabilize WAR results and better reflect true player value, particularly for players who missed significant time during one or more seasons. Weighting was based on the Marcel projection method[^14] and considered three seasons of results, with more recent results being weighted more heavily. For players who have played three or more seasons, the weighted WAR for player-season n was calculated using results from player-seasons n (41.7%), n-1 (33.3%), and n-2 (25%). For players who have played two seasons, the weighted WAR for player-season n was calculated using results from player-seasons n (54.2%), and n-1 (45.8%). For players who have played one season, no adjustments were made.

```{r warning=FALSE}
# Determine a weighted WAR for each player
weight_war <- function(last3_war) {
  player_season <- as.numeric(stringr::str_split_fixed(last3_war, " ", 3))
  if (is.na(player_season[2]))
    player_season[1]
  else if (is.na(player_season[3]))
    weighted.mean(player_season[1:2], c(0.542, 0.458))
  else
    weighted.mean(player_season, c(0.417, 0.333, 0.25))
}

war82_df <- war82_df %>%
  group_by(player) %>%
  arrange(player, season) %>%
  mutate(last3_war = paste(war_82, lag(war_82), lag(war_82, 2))) %>%
  ungroup() %>%
  rowwise() %>%
  mutate(weighted_war_82 = weight_war(last3_war)) 
```

Team names and player names in the war82_df data frame were then re-formatted, in order to facilitate clean joins later in the analysis. The accepted shorthand format for team names (looking at you, Elliote Friedman), and the FIRST.LAST format for player names were used.

```{r warning=FALSE}
# Re-format war82_df team names
team_names_short1 <- c("ANA" = "ANA",
                       "ARI/PHX" = "ARI", 
                       "ARI/PHX" = "PHX",
                       "BOS" = "BOS", 
                       "BUF" = "BUF", 
                       "CGY" = "CGY", 
                       "CAR" = "CAR", 
                       "CHI" = "CHI", 
                       "COL" = "COL", 
                       "CBJ" = "CBJ", 
                       "DAL" = "DAL", 
                       "DET" = "DET", 
                       "EDM" = "EDM", 
                       "FLA" = "FLA", 
                       "LAK" = "L.A",
                       "MIN" = "MIN",
                       "MTL" = "MTL", 
                       "NSH" = "NSH",
                       "NJD" = "N.J",
                       "NYI" = "NYI", 
                       "NYR" = "NYR", 
                       "OTT" = "OTT", 
                       "PHI" = "PHI", 
                       "PIT" = "PIT", 
                       "SJS" = "S.J",
                       "STL" = "STL",
                       "TBL" = "T.B",
                       "TOR" = "TOR",
                       "VAN" = "VAN", 
                       "VGK" = "VGK", 
                       "WSH" = "WSH",
                       "WPG/ATL" = "WPG", 
                       "WPG/ATL" = "ATL")

war82_df$team_one <- names(team_names_short1)[match(war82_df$team_one, team_names_short1)]
war82_df$team_two <- names(team_names_short1)[match(war82_df$team_two, team_names_short1)]
war82_df$team_three <- names(team_names_short1)[match(war82_df$team_three, team_names_short1)]
war82_df$team_four <- names(team_names_short1)[match(war82_df$team_four, team_names_short1)]

# Re-format war82_df player names before separating into first_name, last_name
war82_df$player[war82_df$player == "A.J..GREER"] <- "AJ.GREER"
war82_df$player[war82_df$player == "B.J..CROMBEEN"] <- "BJ.CROMBEEN"
war82_df$player[war82_df$player == "DENIS JR..GAUTHIER"] <- "DENIS JR.GAUTHIER"
war82_df$player[war82_df$player == "J.F..BERUBE"] <- "JF.BERUBE"
war82_df$player[war82_df$player == "J-F.JACQUES"] <- "JS.AUBIN"
war82_df$player[war82_df$player == "J-P.DUMONT"] <- "JP.DUMONT"
war82_df$player[war82_df$player == "J-SEBASTIAN.AUBIN"] <- "JS.AUBIN"
war82_df$player[war82_df$player == "J.T..COMPHER"] <- "JT.COMPHER"
war82_df$player[war82_df$player == "J.T..BROWN"] <- "JT.BROWN"
war82_df$player[war82_df$player == "J.T..MILLER"] <- "JT.MILLER"
war82_df$player[war82_df$player == "MARTIN.ST. LOUIS"] <- "MARTIN.ST LOUIS"
war82_df$player[war82_df$player == "MARTIN.ST. PIERRE"] <- "MARTIN.ST PIERRE"
war82_df$player[war82_df$player == "P.A..PARENTEAU"] <- "PA.PARENTEAU"
war82_df$player[war82_df$player == "PIERRE-ALEX.PARENTEAU"] <- "PA.PARENTEAU"
war82_df$player[war82_df$player == "P.K..SUBBAN"] <- "PK.SUBBAN"
war82_df$player[war82_df$player == "P. J..AXELSSON"] <- "PJ.AXELSSON"
war82_df$player[war82_df$player == "R.J..UMBERGER"] <- "RJ.UMBERGER"
war82_df$player[war82_df$player == "T.J..GALIARDI"] <- "TJ.GALIARDI"
war82_df$player[war82_df$player == "T.J..HENSICK"] <- "TJ.HENSICK"
war82_df$player[war82_df$player == "T.J..OSHIE"] <- "TJ.OSHIE"

# Separate war82_df player names into first_name, last_name
war82_df <- war82_df %>%
  separate(player, c("first_name","last_name"), sep = '\\.') %>%
  # Create first_initials var for join w/ clean_injury_db
  mutate(first_initials = substr(first_name, 1, 2))
```

Required data was selected from the injury_df data frame.

```{r warning=FALSE}
# Create a new injury df with the required data
player_injury_df <- injury_df %>%
  select(chip = Chip, 
         cap_hit = `Cap Hit`, 
         games_missed_injury = `Games Missed`, 
         injury_type = `Injury Type`, 
         player = Player, 
         position = Position, 
         season = Season, 
         team = Team, 
         -`Number of Records`)
```

Team names, player names, and season dates in the player_injury_df data frame were then re-formatted. Player position was also re-formatted, to account for players who are no longer active. Injuries were aggregated on a player-season basis.

```{r warning=FALSE}
# Re-format player_injury_df team names
team_names_short2 <- c("ANA" = "Anaheim", 
                      "ARI/PHX" = "Arizona/Phoenix", 
                      "BOS" = "Boston", 
                      "BUF" = "Buffalo", 
                      "CGY" = "Calgary", 
                      "CAR" = "Carolina", 
                      "CHI" = "Chicago", 
                      "COL" = "Colorado", 
                      "CBJ" = "Columbus", 
                      "DAL" = "Dallas", 
                      "DET" = "Detroit", 
                      "EDM" = "Edmonton", 
                      "FLA" = "Florida", 
                      "LAK" = "Los Angeles", 
                      "MIN" = "Minnesota", 
                      "MTL" = "Montreal", 
                      "NSH" = "Nashville", 
                      "NJD" = "New Jersey", 
                      "NYI" = "NY Islanders", 
                      "NYR" = "NY Rangers", 
                      "OTT" = "Ottawa", 
                      "PHI" = "Philadelphia", 
                      "PIT" = "Pittsburgh", 
                      "SJS" = "San Jose", 
                      "STL" = "St. Louis", 
                      "TBL" = "Tampa Bay", 
                      "TOR" = "Toronto", 
                      "VAN" = "Vancouver", 
                      "VGK" = "Vegas", 
                      "WSH" = "Washington", 
                      "WPG/ATL" = "Winnipeg/Atlanta")

player_injury_df$team <- names(team_names_short2)[match(player_injury_df$team, team_names_short2)]

# Re-format player_injury_df player names before separating into first_name, last_name
player_injury_df$player[player_injury_df$player == "Crombeen, B.J."] <- "Crombeen, BJ"
player_injury_df$player[player_injury_df$player == "King, D.J."] <- "King, DJ"
player_injury_df$player[player_injury_df$player == "Lipon, J.C."] <- "Lipon, JC"
player_injury_df$player[player_injury_df$player == "Dumont, J.P."] <- "Dumont, JP"
player_injury_df$player[player_injury_df$player == "Brown, J.T."] <- "Brown, JT"
player_injury_df$player[player_injury_df$player == "Compher, J.T."] <- "Compher, JT"
player_injury_df$player[player_injury_df$player == "Miller, J.T."] <- "Miller, JT"
player_injury_df$player[player_injury_df$player == "Wyman, J.T."] <- "Wyman, JT"
player_injury_df$player[player_injury_df$player == "St. Louis, Martin"] <- "St Louis, Martin"
player_injury_df$player[player_injury_df$player == "Parenteau, P.A."] <- "Parenteau, PA"
player_injury_df$player[player_injury_df$player == "Subban, P.K."] <- "Subban, PK"
player_injury_df$player[player_injury_df$player == "Umberger, R.J."] <- "Umberger, RJ"
player_injury_df$player[player_injury_df$player == "Brennan, T.J."] <- "Brennan, TJ"
player_injury_df$player[player_injury_df$player == "Brodie, T.J."] <- "Brodie, TJ"
player_injury_df$player[player_injury_df$player == "Galiardi, T.J."] <- "Galiardi, TJ"
player_injury_df$player[player_injury_df$player == "Oshie, T.J."] <- "Oshie, TJ"

player_injury_df <- player_injury_df %>%  
separate(player, c("last_name","first_name"), sep = ', ') %>%
  mutate(last_name = str_to_upper(last_name),
         first_name = str_to_upper(first_name),
         first_initials = substr(first_name, 1, 2)) %>%
  # Re-format season
  separate(season, c("start_year","end_year"), sep = '/') %>%
  mutate(season = paste0(start_year, '-20', end_year)) %>%
  # Re-format position; split into position and status; blanks accounts for weird formatting issue where there isn't an NA
  separate(position, c("position_new", "status", "blanks"), sep = '"') %>%
  # Aggregate injuries on a per-season basis 
  group_by(first_name, 
           first_initials, 
           last_name, 
           team, 
           season, 
           status, 
           position_new, 
           cap_hit) %>%
  summarise(total_games_missed_injury = sum(games_missed_injury), 
            total_chip = sum(chip)) %>%
  ungroup()
```

Required data was selected from the team_stats_df data frame.

```{r warning=FALSE}
# Create a new team stats df with the required data 
team_level_statistics_df <- team_stats_df %>%
  select(team = `X2`, 
         season = `Season`, 
         avg_age = `AvAge`, 
         points = `PTS`, 
         points_pct = `PTS%`) 

# Re-format team_level_statistics_df team names
team_names_short3 <- c("ANA" = "Anaheim Ducks", 
                       "ARI/PHX" = "Arizona Coyotes", 
                       "ARI/PHX" = "Phoenix Coyotes", 
                       "BOS" = "Boston Bruins", 
                       "BUF" = "Buffalo Sabres", 
                       "CGY" = "Calgary Flames", 
                       "CAR" = "Carolina Hurricanes", 
                       "CHI" = "Chicago Blackhawks", 
                       "COL" = "Colorado Avalanche", 
                       "CBJ" = "Columbus Blue Jackets", 
                       "DAL" = "Dallas Stars", 
                       "DET" = "Detroit Red Wings", 
                       "EDM" = "Edmonton Oilers", 
                       "FLA" = "Florida Panthers", 
                       "LAK" = "Los Angeles Kings", 
                       "MIN" = "Minnesota Wild",
                       "MTL" = "Montreal Canadiens", 
                       "NSH" = "Nashville Predators", 
                       "NJD" = "New Jersey Devils", 
                       "NYI" = "New York Islanders", 
                       "NYR" = "New York Rangers", 
                       "OTT" = "Ottawa Senators", 
                       "PHI" = "Philadelphia Flyers", 
                       "PIT" = "Pittsburgh Penguins", 
                       "SJS" = "San Jose Sharks", 
                       "STL" = "St. Louis Blues", 
                       "TBL" = "Tampa Bay Lightning", 
                       "TOR" = "Toronto Maple Leafs", 
                       "VAN" = "Vancouver Canucks", 
                       "VGK" = "Vegas Golden Knights", 
                       "WSH" = "Washington Capitals", 
                       "WPG/ATL" = "Winnipeg Jets", 
                       "WPG/ATL" = "Atlanta Thrashers")

team_level_statistics_df$team <- names(team_names_short3)[match(team_level_statistics_df$team, team_names_short3)]

team_level_statistics_df <- team_level_statistics_df %>%
  arrange(team, season)
```

Retired players were then filtered from the player_injury_df data frame, which was joined to the war82_df data frame to create the injury_value_df data frame.

```{r warning=FALSE}
# Filter retired players, create injury value df
player_injury_df$status[is.na(player_injury_df$status)] <- "active"
player_injury_df <- player_injury_df %>% 
  filter(status == "active")

injury_value_df <- player_injury_df %>%
  left_join(war82_df, 
            by = c("first_initials", "last_name", "season"))
```

Players who were injured but didn't have a WAR record were then filtered. Perry's cut-off for WAR computation was 200 minutes of ice-time. These filtered players were manually examined to confirm that there weren't any formatting issues causing a player to be missing a WAR record.

```{r warning=FALSE}
# Filter players who don't have a WAR record because they didn't meet the 200 minutes  cut-off
injury_value_df <- injury_value_df %>% 
  filter(!is.na(war_82)) %>%
  select(first_name = first_name.x, 
         last_name, 
         position_new, 
         season,
         team, 
         team_one, 
         team_two, 
         team_three, 
         team_four,
         games_played, 
         games_missed, 
         total_games_missed_injury, 
         cap_hit, 
         total_chip, 
         toi, 
         war, 
         war_82, 
         weighted_war_82)
```

The number of wins lost for each player-seaon due to injury was then calculated, with goaltender results being adjusted for expected workload.

A weighted games played value was calculated in order to minimize season-to-season variance, and better reflect true goaltender workload. Weighting was again based on the Marcel projection method. Goaltenders who played more than 41 games on average over the previous three seasons (xth percentile) were expected to play 2/3 of games they missed due to injury (xth percentile), and as such their war_lost value was multiplied by a factor of 0.676. Goaltenders who played fewer than 41 games on average over the previous three seasons were expected to play 1/3 of games they missed due to injury, and as such their war_lost value was multiplied by a factor of 0.324.

This method isn't perfect. For example, it diminishes the impact of former back-ups who were acquired to be a started, and got hurt with their new team, as in the case of Antti Raanta 17/18. As mentioned previously, Perry's WAR data also includes playoff games, which inflated game played totals.

```{r warning=FALSE}
# Determine WAR lost for each player season
player_injury_value_df <- injury_value_df %>%
  mutate(war_GP = weighted_war_82 / 82,
         war_lost = total_games_missed_injury * war_GP)

# Determine a weighted games_played for each player
weighted_gp <- function(last3_gp) {
  player_season <- as.numeric(stringr::str_split_fixed(last3_gp, " ", 3))
  if (is.na(player_season[2]))
    player_season[1]
  else if (is.na(player_season[3]))
    weighted.mean(player_season[1:2], c(0.542, 0.458))
  else
    weighted.mean(player_season, c(0.417, 0.333, 0.25))
}

player_injury_value_df <- player_injury_value_df %>%
  group_by(first_name, last_name) %>%
  arrange(first_name, last_name, season) %>%
  mutate(last3_gp = paste(games_played, lag(games_played), lag(games_played, 2))) %>%
  ungroup() %>%
  rowwise() %>%
  mutate(weighted_games_played = weighted_gp(last3_gp))

player_injury_value_df <- player_injury_value_df %>% 
  mutate(war_lost = case_when(position_new == "G" & weighted_games_played <= 41 ~ 0.324 * war_lost,
                              position_new == "G" & weighted_games_played > 41 ~ 0.676 * war_lost,
                              T ~ war_lost))
```

Results were then compiled, on a team-season basis.

```{r warning=FALSE}
# Attribute injuries to correct teams
team_injury_value_df <- player_injury_value_df %>%
  group_by(team, season) %>%
  summarise(team_war_lost = sum(war_lost),
            team_chip_lost = sum(total_chip),
            team_mgl_lost = sum(total_games_missed_injury))

# Create team_performance_df for analysis of relationship between change in value lost to injury, and change in team performance
team_performance_df <- team_injury_value_df %>%
  left_join(team_level_statistics_df, by = c("team", "season")) 

team_performance_change_df <- team_performance_df %>%
  mutate(last_team_war_lost = lag(team_war_lost),
         team_war_change = team_war_lost - lag(team_war_lost),
         team_chip_change = team_chip_lost - lag(team_chip_lost),
         team_mgl_change = team_mgl_lost - lag(team_mgl_lost),
         points_pct_change = points_pct - lag(points_pct)) %>%
  filter(!is.na(points_pct_change))

# Determine team WAR totals for analysis of relationship between team WAR and points percentage, as referenced in introduction
team_war_relationship <- war82_df %>%
  select(team = team_one, 
         season, 
         war) %>%
  group_by(team, season) %>%
  summarise(team_war = sum(war)) %>%
  left_join(team_level_statistics_df, by = c("team", "season")) %>%
  filter(!is.na(points_pct))
```

### Results

```{r warning=FALSE, message=FALSE, fig.align='center'}
player_injury_value_df %>%
  select(first_name, 
         last_name, 
         position = position_new, 
         season, 
         team, 
         war_lost, 
         games_missed_during_injury = total_games_missed_injury) %>%
  arrange(desc(war_lost)) %>%
  head(20) %>%
  print()
```

```{r warning=FALSE, message=FALSE, fig.align='center'}
team_performance_df %>%
  select(team, 
         season, 
         team_war_lost, 
         team_chip_lost, 
         team_mgl_lost) %>%
  arrange(desc(team_war_lost)) %>%
  head(20) %>%
  print()
```

```{r warning=FALSE, message=FALSE, fig.align='center'}
team_war_relationship_model <- lm(points_pct ~ team_war, team_war_relationship)
team_war_relationship_model_2 <- rlm(points_pct ~ team_war, team_war_relationship, method = "MM")
team_war_relationship_model_3 <- lmrob(points_pct ~ team_war, team_war_relationship)

team_war_relationship_plot <- team_war_relationship %>%
  ggplot(aes(team_war, points_pct)) +
  geom_point() +
  stat_smooth(method = "lm") +
  theme_fivethirtyeight() +
  scale_colour_fivethirtyeight() +
  theme(axis.title = element_text()) + labs(x = "Team WAR", y = "Points Percentage") +
  theme(axis.title.y = element_text(margin = margin(r = 10))) +
  theme(axis.title.x = element_text(margin = margin(t = 10)))

print(team_war_relationship_plot)
summary(team_war_relationship_model)
summary(team_war_relationship_model_2)
summary(team_war_relationship_model_3)
```

Robust regression is an alternative to least squares regression when data are contaminated with outliers or influential observations, and it can also be used for the purpose of detecting influential observations. In instances where we're evaluating the relationship between Team WAR lost and another variable, PIT 2011-2012, PIT 2010-2011, and MTL 2015-2016 are possibly problermatic to an ordinary least squares model, and robust regression is a suitable method.

```{r warning=FALSE, message=FALSE, fig.align='center'}
# https://stats.idre.ucla.edu/r/dae/robust-regression/
# https://stackoverflow.com/questions/17487418/mm-robust-estimation-in-ggplot2-using-stat-smooth-with-method-rlm
# https://stats.stackexchange.com/questions/83826/is-a-weighted-r2-in-robust-linear-model-meaningful-for-goodness-of-fit-analysinsta

avg_age_relationship_model <- rlm(team_war_lost ~ avg_age, team_performance_df, method = "MM")
avg_age_relationship_model_2 <- lmrob(team_war_lost ~ avg_age, team_performance_df)

avg_age_relationship_plot <- team_performance_df %>%
  ggplot(aes(avg_age, team_war_lost)) +
  geom_point() +
  stat_smooth(method = function(formula, 
                                data, 
                                weights = weight) rlm(formula, 
                                                      data, 
                                                      weights = weight, 
                                                      method = "MM"), 
              fullrange = TRUE) +
  theme_fivethirtyeight() +
  scale_colour_fivethirtyeight() +
  theme(axis.title = element_text()) + 
  labs(x = "Average Team Age", y = "WAR Lost to Injury") +
  theme(axis.title.y = element_text(margin = margin(r = 10))) +
  theme(axis.title.x = element_text(margin = margin(t = 10)))

print(avg_age_relationship_plot)
summary(avg_age_relationship_model)
summary(avg_age_relationship_model_2)
```

```{r warning=FALSE, message=FALSE, fig.align='center'}
last_war_lost_model <- rlm(team_war_lost ~ last_team_war_lost, team_performance_change_df, method = "MM")
last_war_lost_model_2 <- lmrob(team_war_lost ~ last_team_war_lost, team_performance_change_df)

last_war_lost_plot <- team_performance_change_df %>%
  ggplot(aes(last_team_war_lost, team_war_lost)) +
  geom_point() +
  stat_smooth(method = function(formula, 
                                data, 
                                weights = weight) rlm(formula, 
                                                      data, 
                                                      weights = weight, 
                                                      method = "MM"), 
              fullrange = TRUE) +
  theme_fivethirtyeight() +
  scale_colour_fivethirtyeight() +
  theme(axis.title = element_text()) + 
  labs(x = "Team WAR Lost n-1", y = "Team WAR Lost n") +
  theme(axis.title.y = element_text(margin = margin(r = 10))) +
  theme(axis.title.x = element_text(margin = margin(t = 10)))

print(last_war_lost_plot)
summary(last_war_lost_model)
summary(last_war_lost_model_2)
```

```{r warning=FALSE, message=FALSE, fig.align='center'}
mgl_lost_relationship_model <- lm(points_pct_change ~ team_mgl_change, team_performance_change_df)

mgl_lost_relationship_plot <- team_performance_change_df %>%
  ggplot(aes(team_mgl_change, points_pct_change)) +
  geom_point() +
  stat_smooth(method = "lm") +
  theme_fivethirtyeight() +
  scale_colour_fivethirtyeight() +
  theme(axis.title = element_text()) + 
  labs(x = "Change in Team MGL", y = "Change in Points Percentage") +
  theme(axis.title.y = element_text(margin = margin(r = 10))) +
  theme(axis.title.x = element_text(margin = margin(t = 10)))

print(mgl_lost_relationship_plot)
summary(mgl_lost_relationship_model)
```


```{r warning=FALSE, message=FALSE, fig.align='center'}
chip_lost_relationship_model <- lm(points_pct_change ~ team_chip_change, team_performance_change_df)

chip_lost_relationship_plot <- team_performance_change_df %>%
  ggplot(aes(team_chip_change, points_pct_change)) +
  geom_point() +
  stat_smooth(method = "lm") +
  theme_fivethirtyeight() +
  scale_colour_fivethirtyeight() +
  theme(axis.title = element_text()) + 
  labs(x = "Change in Team CHIP", y = "Change in Points Percentage") +
  theme(axis.title.y = element_text(margin = margin(r = 10))) +
  theme(axis.title.x = element_text(margin = margin(t = 10)))

print(chip_lost_relationship_plot)
summary(chip_lost_relationship_model)
```

```{r warning=FALSE, message=FALSE, fig.align='center'}
war_lost_relationship_model <- lm(points_pct_change ~ team_war_change, team_performance_change_df)

war_lost_relationship_plot <- team_performance_change_df %>%
  ggplot(aes(team_war_change, points_pct_change)) +
  geom_point() +
  stat_smooth(method = "lm") +
  theme_fivethirtyeight() +
  scale_colour_fivethirtyeight() +
  theme(axis.title = element_text()) + 
  labs(x = "Change in Team WAR Lost", y = "Change in Points Percentage") +
  theme(axis.title.y = element_text(margin = margin(r = 10))) +
  theme(axis.title.x = element_text(margin = margin(t = 10)))

print(war_lost_relationship_plot)
summary(war_lost_relationship_model)
```

Interestingly, there does not appear to be a strong relationship between the change value lost to injury, and team points percentage.

Age vs. WAR lost

### Discussion

Aaron Kneeland’s research found no consistent correlation between travel, rest periods, and man games lost, though it should be noted that his study only examined two seasons of data (2014). 

This analysis is also limited by its exclusion of goaltenders. This was purposeful, given that the demands of the positions differ significantly, and injuries tend to be correspondingly different. However, goalie injuries are clearly impactful, and goaltender workload is a large, large issue that needs solving. 

https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0201264
https://bjsm.bmj.com/content/50/5/273
https://channels.theinnovationenterprise.com/interviews/james-bunce-head-of-sport-science-premier-league
https://www.forbes.com/sites/stevemccaskill/2018/05/11/how-sports-science-helps-newcastle-united-prevent-injury-and-identify-academy-stars/#3e0539053b8c


### Acknowledgements 

Shout outs to: 

* Matt Barlowe, Matt Cane, Domenic Galamini Jr., Alex Novet, and Josh and Luke Younggren for graciously and patiently answering my questions at various points over the past year. 
* Daniel Meyers, who provided feedback on ____.
* Daniel Jenkins, Matt McConnell, and Jane Vandale, who were early reviewers.
* Manny Perry, Hockey Reference, and NHL Injury Viz for their respective public work that provided the basis for this paper; hockey analytics is better for your efforts.

[^1]: NHL Injury Viz, “NHL Injury Viz: Explainer,” http://nhlinjuryviz.blogspot.com/2016/12/explainer.html, (December 28, 2016).
[^2]: Alan Ryder, “Player Contribution: A method for allocating credit for a team’s performance to the individual contributors on a hockey team,” http://www.hockeyanalytics.com/Research_files/Player_Contribution_System.pdf, (August 2014).
[^3]: Tom Awad, “Numbers On Ice: Understanding GVT, Part 1,” http://www.hockeyprospectus.com/puck/article.php?articleid=233, (July 30, 2009).
[^4]: Brian MacDonald, “Adjusted Plus-Minus for NHL Players using Ridge Regression with Goals, Shots, Fenwick, and Corsi,” Journal of Quantitative Analysis in Sports Vol. 8, Issue 3 (October 2012).
[^5]: Michael Schuckers and James Curro, “Total Hockey Rating (THoR): A comprehensive statistical rating of National Hockey League forwards and defensemen based upon all on-ice events,” http://statsportsconsulting.com/main/wp-content/uploads/Schuckers_Curro_MIT_Sloan_THoR.pdf, (March 1, 2013).
[^6]: Robert B. Gramacy, Matthew A. Taddy, and Shane T. Jensen, “Estimating Player Contribution in Hockey with Regularized Logistic Regression,” https://arxiv.org/abs/1209.5026, (Jan 12, 2013).
[^7]: A.C. Thomas et al., “Competing process hazard function models for player ratings in ice hockey,”, The Annals of Applied Statistics Vol. 7, No. 3 (2013), 1497-1524.
[^8]: Emmanuel Perry, “Wins Above Replacement,” http://www.corsica.hockey/misc/war_notebook.html, (April 21, 2018).
[^9]: A.C. Thomas, “The Road to WAR Series: Index,” http://blog.war-on-ice.com/index.html%3Fp=429.html, (October 5, 2014).
[^10]: Dawson Sprigings, “A Primer on @DTMAboutHeart’s WAR Model,” https://hockey-graphs.com/2016/10/24/a-primer-on-dtmabouthearts-war-model/, (October 24, 2016).
[^11]: Emmanuel Perry, “The Art of WAR,” http://www.corsica.hockey/blog/2017/05/20/the-art-of-war/, (May 20, 2017).
[^12]: Josh Younggren and Luke Younggren, "Wins Above Replacement: Replacement Level, Decisions, Results, and Final Remarks," https://hockey-graphs.com/2019/01/18/wins-above-replacement-replacement-level-decisions-results-and-final-remarks-part-3/, (January 18, 2019)
[^13]: Cam Lawrence, "How to build a contender – Part 1: WAR, what is it good for?," https://canucksarmy.com/2015/08/17/how-to-build-a-contender-part-1-war-what-is-it-good-for/, (August 17, 2015)
[^14]: Henry Druschel, "A guide to the projection systems," https://www.beyondtheboxscore.com/2016/2/22/11079186/projections-marcel-pecota-zips-steamer-explained-guide-math-is-fun, (February 22, 2016)