---
title: 'Winning the battle, but losing the WAR: Examining the value of skater injuries
  in the NHL'
author: "Luke Steer (@lukeasteer)"
date: "April 1, 2019"
output:
  html_document:
    theme: default
    highlight: default
---

### "All players are equal, but some players are more equal than others""

Not all player injuries in the NHL are equivalent, which is a concept so intuitive that it hardly bears explaining. For example, you’d be hard-pressed to find an analyst or fan who believes losing Iiro Pakarinen for four weeks is as impactful as losing Connor McDavid. However, the discussion of injuries in the NHL is most often centred around Man Games Lost (MGL), a basic statistic which limits the quality of analysis and discussion. More modern statistics such as Cap Hit of Injured Player (CHIP)[^1] and Time-On-Ice of Injured Player (TOIIP) attempt to account for the value of the player that was lost to injury, but fall short, given that salary is influenced by player seniority and market factors that vary on a year-to-year basis, while time-on-ice is influenced by coaching staff bias and decision-making. 

Since the origin of hockey analytics, there have been attempts to develop single-value evaluation metrics, useful for comparison of total player value across different teams and seasons. Early examples include Ryder’s Player Contribution (2003)[^2], Awad’s Goals Versus Threshold (2009)[^3], MacDonald’s Adjusted Plus-Minus (2012)[^4], Schuckers and Curro’s Total Hockey Ratings (2013)[^5], Gramacy, Taddy, and Jensen’s Logistic Event Ratings (2013)[^6], and Thomas et al’s Mean Even Strength Hazard Model (2013)[^7]. 

WAR, first developed by sabermetricians in baseball circles, is an evaluative metric intended to approximate a player’s impact in units of wins added, measured against what a replacement level player would be expected to offer in the same circumstances[^8]. Recently, this framework has been applied in the context of the NHL by Thomas et al. (2014)[^9], Sprigings (2016)[^10], Perry (2017)[^11], and Solberg and Solberg (2018)[^12]. For the purposes of this evaluation, I have used Perry's version of WAR as a measurement of player value. Replacement level is not uniquely defined, but it is commonly equated to players earning a league-minimum salary, and Perry's WAR adopts this definition.

[insert figure]

In this paper, I posit that player value should be considered when evaluating injuries. Team strength, in terms of player WAR, and team points percentage were highly correlated for the seasons examined in this study (Adjusted R-squared = 0.6576). This finding is supported by Cam Lawrence's 2015 study[^13], which found that team strength, in terms of Spriging's Goals Above Replacement, was highly correlated with team points (R2 = 0.7231) for seasons between 2009-2010 and 2014-2015. As such, it stands to reason that losing stronger players would have an impact on team points percentage.

Furthermore, accounting for value lost to injury enables analysis of medical staff efficacy. A team’s most productive players tend to play the most, and as such are subjected to a higher degree of injury risk, both in terms of overuse injuries and exposure to deleterious on-ice events such as hits and blocked shots. However, it is common for elite players to receive special or preferential medical treatment that should, in theory, mitigate their injury risk to some extent. An efficient medical staff is one that is able to identify risk factors and reduce training load and exposure prior to injury, while also helping players who have been injured return to action more successfully, ultimately improving the player's availability, performance, and total output. In developing more nuanced metrics for evaluating value lost to injury, we’re able to more accurately identify teams with recurring injury problems, which is the first step towards developing appropriate solutions to these problems.

### Method

I examined 9 seasons of data, from the 2009-2010 season, through the 2017-2018 season. Injury data is sourced from nhlinjuryviz.blogspot.com, and WAR data is sourced from corsica.hockey. 

Assumptions:

* Perry does not differentiate between the regular season and playoffs when calculating WAR; all games are incorporated into a single value for the season. I’ve assumed that a) on aggregate, player performance is relatively stable between the regular season and the playoffs, and b) WAR is more reflective of regular season performance, due to the comparatively smaller number of playoff games in the sample. For these reasons, I’ve chosen not to undergo the labour-intensive process of re-calculating WAR/82.
* In cases where a player missed an entire season due to injury, I’ve taken the average WAR of the seasons preceding and following the missed season. 
* In cases where a player missed two consecutive seasons, or missed significant portions of two or more seasons in a row (i.e. < 20 GP), while still being considered active, I’ve taken the average WAR/82 of the seasons preceding and following the missed season.
* In cases where a player missed two consecutive seasons, or missed significant portions of two or more seasons in a row (i.e. < 20 GP), while still being considered active, and then proceeded to retire, I’ve used a Marcel projection to determine their WAR/82 for the missed seasons. 
* There are no cases where a player missed three consecutive seasons,  or missed significant portions of three or more seasons in a row (i.e. < 20 GP), while still being considered active. 
* The quality of the replacement player is not considered; my aim is not to reward deep organizations, nor to penalize shallow ones; this is strictly an injury evaluation.

```{r warning=FALSE, message=FALSE}
# Install and load packages
library(tidyverse)

# Load WAR data, injury data 
war_df <- read_csv("war_ratings_2019-02-14.csv")
injury_df <- read_csv("NHL_Injury_Database_data.csv")
team_stats_df <- read_csv("NHL_team_data.csv")
```

```{r warning=FALSE}
# Create a new WAR df with the required data 
war82_df <- war_df %>%
  dplyr::select(player = Player, season = Season, team = Team, games_played = GP, toi = TOI, war_82 = `WAR/82`, war = WAR) %>%
  # Account for mid-season trades
  separate(team, c("team_one","team_two", "team_three", "team_four"), sep = '/') %>%
  # Create games_missed var, for comparison to games_missed_injury; account for absences for personal reasons, etc.
  mutate(games_missed = 82 - games_played) 

# Determine a weighted WAR for each player
weight_war <- function(last3_war) {
  player_season <- as.numeric(stringr::str_split_fixed(last3_war, " ", 3))
  if (is.na(player_season[2]))
    player_season[1]
  else if (is.na(player_season[3]))
    weighted.mean(player_season[1:2], c(0.542, 0.458))
  else
    weighted.mean(player_season, c(0.417, 0.333, 0.25))
}

war82_df <- war82_df %>%
  group_by(player) %>%
  arrange(player, season) %>%
  mutate(last3_war = paste(war_82, lag(war_82), lag(war_82, 2))) %>%
  ungroup() %>%
  rowwise() %>%
  mutate(weighted_war_82 = weight_war(last3_war)) 

# Re-format war82_df team names
team_names_short1 <- c("ANA" = "ANA",
                       "ARI/PHX" = "ARI", 
                       "ARI/PHX" = "PHX",
                       "BOS" = "BOS", 
                       "BUF" = "BUF", 
                       "CGY" = "CGY", 
                       "CAR" = "CAR", 
                       "CHI" = "CHI", 
                       "COL" = "COL", 
                       "CBJ" = "CBJ", 
                       "DAL" = "DAL", 
                       "DET" = "DET", 
                       "EDM" = "EDM", 
                       "FLA" = "FLA", 
                       "LAK" = "L.A",
                       "MIN" = "MIN",
                       "MTL" = "MTL", 
                       "NSH" = "NSH",
                       "NJD" = "N.J",
                       "NYI" = "NYI", 
                       "NYR" = "NYR", 
                       "OTT" = "OTT", 
                       "PHI" = "PHI", 
                       "PIT" = "PIT", 
                       "SJS" = "S.J",
                       "STL" = "STL",
                       "TBL" = "T.B",
                       "TOR" = "TOR",
                       "VAN" = "VAN", 
                       "VGK" = "VGK", 
                       "WSH" = "WSH",
                       "WPG/ATL" = "WPG", 
                       "WPG/ATL" = "ATL")

war82_df$team_one <- names(team_names_short1)[match(war82_df$team_one, team_names_short1)]
war82_df$team_two <- names(team_names_short1)[match(war82_df$team_two, team_names_short1)]
war82_df$team_three <- names(team_names_short1)[match(war82_df$team_three, team_names_short1)]
war82_df$team_four <- names(team_names_short1)[match(war82_df$team_four, team_names_short1)]

# Re-format war82_df player names before separating into first_name, last_name
war82_df$player[war82_df$player == "A.J..GREER"] <- "AJ.GREER"
war82_df$player[war82_df$player == "B.J..CROMBEEN"] <- "BJ.CROMBEEN"
war82_df$player[war82_df$player == "DENIS JR..GAUTHIER"] <- "DENIS JR.GAUTHIER"
war82_df$player[war82_df$player == "J.F..BERUBE"] <- "JF.BERUBE"
war82_df$player[war82_df$player == "J-F.JACQUES"] <- "JS.AUBIN"
war82_df$player[war82_df$player == "J-P.DUMONT"] <- "JP.DUMONT"
war82_df$player[war82_df$player == "J-SEBASTIAN.AUBIN"] <- "JS.AUBIN"
war82_df$player[war82_df$player == "J.T..COMPHER"] <- "JT.COMPHER"
war82_df$player[war82_df$player == "J.T..BROWN"] <- "JT.BROWN"
war82_df$player[war82_df$player == "J.T..MILLER"] <- "JT.MILLER"
war82_df$player[war82_df$player == "MARTIN.ST. LOUIS"] <- "MARTIN.ST LOUIS"
war82_df$player[war82_df$player == "MARTIN.ST. PIERRE"] <- "MARTIN.ST PIERRE"
war82_df$player[war82_df$player == "P.A..PARENTEAU"] <- "PA.PARENTEAU"
war82_df$player[war82_df$player == "PIERRE-ALEX.PARENTEAU"] <- "PA.PARENTEAU"
war82_df$player[war82_df$player == "P.K..SUBBAN"] <- "PK.SUBBAN"
war82_df$player[war82_df$player == "P. J..AXELSSON"] <- "PJ.AXELSSON"
war82_df$player[war82_df$player == "R.J..UMBERGER"] <- "RJ.UMBERGER"
war82_df$player[war82_df$player == "T.J..GALIARDI"] <- "TJ.GALIARDI"
war82_df$player[war82_df$player == "T.J..HENSICK"] <- "TJ.HENSICK"
war82_df$player[war82_df$player == "T.J..OSHIE"] <- "TJ.OSHIE"

# Separate war82_df player names into first_name, last_name
war82_df <- war82_df %>%
  separate(player, c("first_name","last_name"), sep = '\\.') %>%
  # Create first_initials var for join w/ clean_injury_db
  mutate(first_initials = substr(first_name, 1, 2))
```

```{r warning=FALSE}
# Create a new injury df with the required data
player_injury_df <- injury_df %>%
  dplyr::select(chip = Chip, cap_hit = `Cap Hit`, games_missed_injury = `Games Missed`, injury_type = `Injury Type`, 
         player = Player, position = Position, season = Season, team = Team, -`Number of Records`)

# Re-format player_injury_df team names
team_names_short2 <- c("ANA" = "Anaheim", 
                      "ARI/PHX" = "Arizona/Phoenix", 
                      "BOS" = "Boston", 
                      "BUF" = "Buffalo", 
                      "CGY" = "Calgary", 
                      "CAR" = "Carolina", 
                      "CHI" = "Chicago", 
                      "COL" = "Colorado", 
                      "CBJ" = "Columbus", 
                      "DAL" = "Dallas", 
                      "DET" = "Detroit", 
                      "EDM" = "Edmonton", 
                      "FLA" = "Florida", 
                      "LAK" = "Los Angeles", 
                      "MIN" = "Minnesota", 
                      "MTL" = "Montreal", 
                      "NSH" = "Nashville", 
                      "NJD" = "New Jersey", 
                      "NYI" = "NY Islanders", 
                      "NYR" = "NY Rangers", 
                      "OTT" = "Ottawa", 
                      "PHI" = "Philadelphia", 
                      "PIT" = "Pittsburgh", 
                      "SJS" = "San Jose", 
                      "STL" = "St. Louis", 
                      "TBL" = "Tampa Bay", 
                      "TOR" = "Toronto", 
                      "VAN" = "Vancouver", 
                      "VGK" = "Vegas", 
                      "WSH" = "Washington", 
                      "WPG/ATL" = "Winnipeg/Atlanta")

player_injury_df$team <- names(team_names_short2)[match(player_injury_df$team, team_names_short2)]

# Re-format player_injury_df player names before separating into first_name, last_name
player_injury_df$player[player_injury_df$player == "Crombeen, B.J."] <- "Crombeen, BJ"
player_injury_df$player[player_injury_df$player == "King, D.J."] <- "King, DJ"
player_injury_df$player[player_injury_df$player == "Lipon, J.C."] <- "Lipon, JC"
player_injury_df$player[player_injury_df$player == "Dumont, J.P."] <- "Dumont, JP"
player_injury_df$player[player_injury_df$player == "Brown, J.T."] <- "Brown, JT"
player_injury_df$player[player_injury_df$player == "Compher, J.T."] <- "Compher, JT"
player_injury_df$player[player_injury_df$player == "Miller, J.T."] <- "Miller, JT"
player_injury_df$player[player_injury_df$player == "Wyman, J.T."] <- "Wyman, JT"
player_injury_df$player[player_injury_df$player == "Parenteau, P.A."] <- "Parenteau, PA"
player_injury_df$player[player_injury_df$player == "Subban, P.K."] <- "Subban, PK"
player_injury_df$player[player_injury_df$player == "Umberger, R.J."] <- "Umberger, RJ"
player_injury_df$player[player_injury_df$player == "Brennan, T.J."] <- "Brennan, TJ"
player_injury_df$player[player_injury_df$player == "Brodie, T.J."] <- "Brodie, TJ"
player_injury_df$player[player_injury_df$player == "Galiardi, T.J."] <- "Galiardi, TJ"
player_injury_df$player[player_injury_df$player == "Oshie, T.J."] <- "Oshie, TJ"

player_injury_df <- player_injury_df %>%  
separate(player, c("last_name","first_name"), sep = ', ') %>%
  mutate(last_name = str_to_upper(last_name)) %>%
  mutate(first_name = str_to_upper(first_name)) %>%
  mutate(first_initials = substr(first_name, 1, 2)) %>%
  # Re-format season
  separate(season, c("start_year","end_year"), sep = '/') %>%
  mutate(season = paste0(start_year, '-20', end_year)) %>%
  # Re-format position; split into position and status; blanks accounts for weird formatting issue where there isn't an NA
  separate(position, c("position_new", "status", "blanks"), sep = '"') %>%
  # Aggregate injuries on a per-season basis 
  group_by(first_name, first_initials, last_name, team, season, status, position_new, cap_hit) %>%
  summarise(total_games_missed_injury = sum(games_missed_injury), total_chip = sum(chip)) %>%
  ungroup()
```

```{r warning=FALSE}
# Create a new team stats df with the required data 
team_level_statistics_df <- team_stats_df %>%
  dplyr::select(team = `X2`, season = `Season`, avg_age = `AvAge`, points = `PTS`, points_pct = `PTS%`) 

# Re-format team_level_statistics_df team names
team_names_short3 <- c("ANA" = "Anaheim Ducks", 
                       "ARI/PHX" = "Arizona Coyotes", 
                       "ARI/PHX" = "Phoenix Coyotes", 
                       "BOS" = "Boston Bruins", 
                       "BUF" = "Buffalo Sabres", 
                       "CGY" = "Calgary Flames", 
                       "CAR" = "Carolina Hurricanes", 
                       "CHI" = "Chicago Blackhawks", 
                       "COL" = "Colorado Avalanche", 
                       "CBJ" = "Columbus Blue Jackets", 
                       "DAL" = "Dallas Stars", 
                       "DET" = "Detroit Red Wings", 
                       "EDM" = "Edmonton Oilers", 
                       "FLA" = "Florida Panthers", 
                       "LAK" = "Los Angeles Kings", 
                       "MIN" = "Minnesota Wild",
                       "MTL" = "Montreal canadiens", 
                       "NSH" = "Nashville Predators", 
                       "NJD" = "New Jersey Deviles", 
                       "NYI" = "New York Islanders", 
                       "NYR" = "New York Rangers", 
                       "OTT" = "Ottawa Senators", 
                       "PHI" = "Philadelphia Flyers", 
                       "PIT" = "Pittsburgh Penguins", 
                       "SJS" = "San Jose Sharks", 
                       "STL" = "St. Louis Blues", 
                       "TBL" = "Tampa Bay Lightning", 
                       "TOR" = "Toronto Maple Leafs", 
                       "VAN" = "Vancouver Canucks", 
                       "VGK" = "Vegas Golden Knights", 
                       "WSH" = "Washington Capitals", 
                       "WPG/ATL" = "Winnipeg Jets", 
                       "WPG/ATL" = "Atlanta Thrashers")

team_level_statistics_df$team <- names(team_names_short3)[match(team_level_statistics_df$team, team_names_short3)]

team_level_statistics_df <- team_level_statistics_df %>%
  arrange(team, season)
```

```{r warning=FALSE}
# Filter retired players, goaltenders, create injury value db
player_injury_df$status[is.na(player_injury_df$status)] <- "active"
player_injury_df <- player_injury_df %>% 
  filter(status == "active")

injury_value_df <- player_injury_df %>%
  left_join(war82_df, by = c("first_initials", "last_name", "season"))

# Assume for the time being that players without WAR records aren't relevant (didn't meet cut-off)
injury_value_df <- injury_value_df %>% 
  filter(!is.na(war_82)) %>%
  dplyr::select(first_name = first_name.x, last_name, position_new, season,
           team, team_one, team_two, team_three, team_four,
           games_played, games_missed, total_games_missed_injury, cap_hit, total_chip, toi, war, war_82, weighted_war_82)

# Manually examine players who were traded to determine if injured games were correctly attributed
# STILL NEED TO DO THIS
injury_value_df$team_two[is.na(injury_value_df$team_two)] <- "none"
injury_value_df$team_three[is.na(injury_value_df$team_three)] <- "none"
injury_value_df$team_four[is.na(injury_value_df$team_four)] <- "none"

traded_players <- injury_value_df %>%
  filter(team_two != "none")

# Filter out goalies, because we won't be including those in the analysis
skater_injury_value_df <- injury_value_df %>%
  filter(position_new != "G")

# Determine WAR lost for each player season
skater_injury_value_df <- skater_injury_value_df %>%
  mutate(war_GP = weighted_war_82 / 82) %>%
  mutate(war_lost = total_games_missed_injury * war_GP)

# Attribute injuries to correct teams
team_injury_value_df <- skater_injury_value_df %>%
  group_by(team, season) %>%
  summarise(team_war_lost = sum(war_lost),
            team_chip = sum(total_chip),
            man_games_lost = sum(total_games_missed_injury))

# Create working df for analysis
team_performance_df <- team_injury_value_df %>%
  left_join(team_level_statistics_df, by = c("team", "season")) 

team_performance_change_df <- team_performance_df %>%
  mutate(team_war_change = team_war_lost - lag(team_war_lost)) %>%
  mutate(team_chip_change = team_chip - lag(team_chip)) %>%
  mutate(team_mgl_change = man_games_lost - lag(man_games_lost)) %>%
  mutate(points_pct_change = points_pct - lag(points_pct)) %>%
  filter(!is.na(points_pct_change))
```

### Results

```{r warning=FALSE, message=FALSE}

# Create war df that includes goalie results for comparison to points_pct
team_war_totals <- war82_df %>%
  dplyr::select(team = team_one, season, war) %>%
  group_by(team, season) %>%
  summarise(team_war = sum(war))

team_war_relationship <- team_war_totals %>%
  left_join(team_level_statistics_df, by = c("team", "season")) %>%
   filter(!is.na(points_pct))

war_relations <- lm(points_pct ~ team_war, team_war_relationship)
```

```{r warning=FALSE, message=FALSE}
library(MASS)

model_1 <- rlm(points_pct_change ~ team_war_change, team_performance_change_df)
```

Age vs. WAR lost

### Discussion

Aaron Kneeland’s research found no consistent correlation between travel, rest periods, and man games lost, though it should be noted that his study only examined two seasons of data (2014). 

This analysis is also limited by its exclusion of goaltenders. This was purposeful, given that the demands of the positions differ significantly, and injuries tend to be correspondingly different. However, goalie injuries are clearly impactful, and goaltender workload is a large, large issue that needs solving. 

[^1]: NHL Injury Viz, “NHL Injury Viz: Explainer,” http://nhlinjuryviz.blogspot.com/2016/12/explainer.html, (December 28, 2016).
[^2]: Alan Ryder, “Player Contribution: A method for allocating credit for a team’s performance to the individual contributors on a hockey team,” http://www.hockeyanalytics.com/Research_files/Player_Contribution_System.pdf, (August 2014).
[^3]: Tom Awad, “Numbers On Ice: Understanding GVT, Part 1,” http://www.hockeyprospectus.com/puck/article.php?articleid=233, (July 30, 2009).
[^4]: Brian MacDonald, “Adjusted Plus-Minus for NHL Players using Ridge Regression with Goals, Shots, Fenwick, and Corsi,” Journal of Quantitative Analysis in Sports Vol. 8, Issue 3 (October 2012).
[^5]: Michael Schuckers and James Curro, “Total Hockey Rating (THoR): A comprehensive statistical rating of National Hockey League forwards and defensemen based upon all on-ice events,” http://statsportsconsulting.com/main/wp-content/uploads/Schuckers_Curro_MIT_Sloan_THoR.pdf, (March 1, 2013).
[^6]: Robert B. Gramacy, Matthew A. Taddy, and Shane T. Jensen, “Estimating Player Contribution in Hockey with Regularized Logistic Regression,” https://arxiv.org/abs/1209.5026, (Jan 12, 2013).
[^7]: A.C. Thomas et al., “Competing process hazard function models for player ratings in ice hockey,”, The Annals of Applied Statistics Vol. 7, No. 3 (2013), 1497-1524.
[^8]: Emmanuel Perry, “Wins Above Replacement,” http://www.corsica.hockey/misc/war_notebook.html, (April 21, 2018).
[^9]: A.C. Thomas, “The Road to WAR Series: Index,” http://blog.war-on-ice.com/index.html%3Fp=429.html, (October 5, 2014).
[^10]: Dawson Sprigings, “A Primer on @DTMAboutHeart’s WAR Model,” https://hockey-graphs.com/2016/10/24/a-primer-on-dtmabouthearts-war-model/, (October 24, 2016).
[^11]: Emmanuel Perry, “The Art of WAR,” http://www.corsica.hockey/blog/2017/05/20/the-art-of-war/, (May 20, 2017).
[^12]: Luke Younggren and Josh Younggren, "Wins Above Replacement: Replacement Level, Decisions, Results, and Final Remarks," https://hockey-graphs.com/2019/01/18/wins-above-replacement-replacement-level-decisions-results-and-final-remarks-part-3/, (January 18, 2019)
[^13]: Cam Lawrence, "How to build a contender – Part 1: WAR, what is it good for?"", https://canucksarmy.com/2015/08/17/how-to-build-a-contender-part-1-war-what-is-it-good-for/, (August 17, 2015)